---
title: "Energy monitor"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(ggplot2)
```

## Introduction

Trying to reduce our private consumption of electricity and natural gas I started 
to collect weekly readings from our gas and electricity meter. 

Data is manually collected via a data sheet (`raw/energymonitor.ods`) and then 
visualized using R and `ggplot2`.

## Data preparation

Data is imported from the spreadsheet (here LibreOffice Calc format `ods`) as a 
`data.table` object, the latest (and yet empty) rows of the template are 
removed and only the columns of interest are selected. 

```{r}
rdg = readODS::read_ods(
  "data/energymonitor.ods"
  , sheet = "data"
) |> 
  subset(!is.na(rdg_gas)) |> 
  subset(select = c("date", "rdg_gas", "rdg_pow")) |> 
  as.data.table()

rdg[, date := as.POSIXct(date, format = "%m/%d/%Y")]
rdg[, week := strftime(date, format = "%V")]

head(rdg)
```

The gas meter records the natural gas consumption continuously in 
cubic meters. To convert cubic meters readings to kWh per week one has to 

* calculate the difference of successive readings. 
* multiply the difference with two factors, the calorific value (German: "Brennwert") 
and the volume correction factor (German: "Zustandszahl"). If these are unknown, 
a good estimate can be achieved by multiplying the cubic meters with 10. In my 
case, I took both factors from the last invoice of my provider. 

For electricity, consumption is recorded in kWh, so taking the difference gives 
the consumption per week. 

```{r}
cbm2kwh = 0.9355 * 11.517 

rdg[, use_pow := c(NA_real_, diff(rdg_pow))]
rdg[, use_gas := c(NA_real_, diff(rdg_gas))]
rdg[, use_gas_kwh := use_gas * cbm2kwh]

head(rdg)
```


