---
title: "Energy monitor"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(ggplot2)
```

## Introduction

Trying to reduce our private consumption of electricity and natural gas I started 
to collect weekly readings from our gas and electricity meter. 

Data is manually collected via a data sheet (`raw/energymonitor.ods`) and then 
visualized using R and `ggplot2`.

## Data preparation

Data is imported from the spreadsheet (here LibreOffice Calc format `ods`) as a 
`data.table` object, the latest (and yet empty) rows of the template are 
removed and only the columns of interest are selected. 

```{r}
rdg = readODS::read_ods(
  "data/energymonitor.ods"
  , sheet = "data"
) |> 
  subset(!is.na(rdg_gas)) |> 
  subset(select = c("date", "rdg_gas", "rdg_pow")) |> 
  as.data.table()

rdg[, date := as.POSIXct(date, format = "%m/%d/%Y")]
rdg[, season := year(date)]
rdg[, doy := yday(date)]

rdg[
  , offset_jul := yday(
    as.Date(
      paste(
        year(date)
        , "07"
        , "01"
        , sep = "-"
      )
    )
  )
]

rdg[
  , offset_dec := yday(
    as.Date(
      paste(
        year(date) - 1
        , "12"
        , "31"
        , sep = "-"
      )
    )
  )
]

rdg[, doy := doy - offset_jul + 1L]
rdg[doy <= 0, season := season - 1L]
rdg[doy <= 0, doy := doy + offset_dec]
rdg[, c("offset_jul", "offset_dec") := NULL]

rdg[, lab := substring(date, first = 6)]
```

The gas meter records the natural gas consumption continuously in cubic meters. 
To convert cubic meters readings to kWh per week one has to 

* calculate the difference of successive readings 
* multiply the difference with two factors, the calorific value (German: 
"Brennwert") and the volume correction factor (German: "Zustandszahl"). If 
these are unknown, a good estimate can be achieved by multiplying the cubic 
meters with 10. In my case, I took both factors from the last invoice of my 
provider. 

For electricity, consumption is recorded in kWh, so taking the difference gives 
the consumption per week. 

```{r}
cbm2kwh = 0.9355 * 11.517 

rdg[, use_pow_kwh := c(NA_real_, diff(rdg_pow))]
rdg[, use_gas := c(NA_real_, diff(rdg_gas))]
rdg[, use_gas_kwh := use_gas * cbm2kwh]

rdg[, c("rdg_pow", "rdg_gas", "use_gas") := NULL]

rdg[, cum_pow_kwh := use_pow_kwh]
rdg[, cum_gas_kwh := use_gas_kwh]
rdg[1L, cum_pow_kwh := 0]
rdg[1L, cum_gas_kwh := 0]
rdg[, cum_pow_kwh := cumsum(cum_pow_kwh), by = season]
rdg[, cum_gas_kwh := cumsum(cum_gas_kwh), by = season]

# rdg[, sm_gas_kwh := runmed(use_gas_kwh, k = 7L, endrule = "median")]
# rdg[, sm_pow_kwh := runmed(use_pow_kwh, k = 7L, endrule = "median")]

head(rdg)
```

## Data preparation

I'll do two types of plots: A continuous plot of week-on-week consumption and 
a accumulation per season, both for natural gas and electricity. So, finally 
this ends up in 2*2 plots.

```{r}
rdg = melt(
  rdg
  ,id.vars = c("date", "season", "doy", "lab")
  , measure.vars = c("use_gas_kwh", "use_pow_kwh", "cum_gas_kwh", "cum_pow_kwh")
  , value.name = "use"
  , variable.name = "type"
  , na.rm = TRUE
)
```


```{r}
#| label: plt-gas
ggplot(
  rdg[grep("gas", x = type)]
  ,  mapping = aes(doy, y = use, color = factor(season), group = season)
) + 
  geom_line() + 
  scale_x_continuous("") + 
  scale_y_continuous("Consumption [kWh]") + 
  scale_color_manual(
    ""
    , values = c(
      "2022" = "gray80"
      , "2023" = "gray20"
    )
  ) + 
  facet_wrap(
    "type"
    , ncol = 1
    , scales = "free_y"
    , labeller = as_labeller(
      c(
        use_gas_kwh = "gas [kWh]"
        , cum_gas_kwh = "gas [kWh]"
      )
    )
  ) +
  theme_bw() + 
  theme(legend.position = "none")
```
```{r}
#| label: plt-pow
ggplot(
  rdg[grep("pow", x = type)]
  ,  mapping = aes(doy, y = use, color = factor(season), group = season)
) + 
  geom_line() + 
  scale_x_continuous("") + 
  scale_y_continuous("Consumption [kWh]") + 
  scale_color_manual(
    ""
    , values = c(
      "2022" = "gray80"
      , "2023" = "gray20"
    )
  ) + 
  facet_wrap(
    "type"
    , ncol = 1
    , scales = "free_y"
    , labeller = as_labeller(
      c(
        use_pow_kwh = "pow [kWh]"
        , cum_pow_kwh = "pow [kWh]"
      )
    )
  ) +
  theme_bw() + 
  theme(legend.position = "none")
```

